Interspecies comparisons of Mg/Ca ratios in limpet shells

### 2024-06-27 Manuscript_with_Heshkaia.tex
first draft

\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Palaeo3}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Interspecies comparisons of Mg/Ca ratios in limpet shells},
  pdfauthor={Niklas Hausmann; Donna Surge; Francisco Zangrando; Angelica Tivoli; Ivan Briz-Godino},
  pdfkeywords={Sclerochronology, Limpets, Elemental Ratio, Mg/Ca},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\author[3]{Francisco Zangrando%
%
}

\author[3]{Angelica Tivoli%
%
}

\author[4]{Ivan Briz-Godino%
%
}

\affiliation[1]{organization={Leibniz Zentrum für
Archäologie},addressline={Ludwig-Lindenschmit-Forum
1},city={Mainz},country={Germany},countrysep={,},postcode={55116},postcodesep={}}
\affiliation[2]{organization={University of North
Carolina},addressline={104 South Road, 225 Geology
Building},city={Chapel Hill,
NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
\affiliation[3]{organization={CONICET (Consejo Nacional de
Investigaciones Científicas y Técnicas)},addressline={Avenida Maipú
305},city={Ushuaia},country={Argentina},countrysep={,},postcode={V9410BJA},postcodesep={}}
\affiliation[4]{organization={},country={Spain},countrysep={,},postcodesep={}}

@@

\begin{abstract}
This study provides a short reassessment of the use of Magnesium to
Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea
surface temperatures. While \emph{Patella vulgata} along the Spanish
shoreline has since then repeatedly produced reliable correlations
between sea surface temperature and Mg/Ca ratios, this is not the case
for other patelloid species. \emph{Patella vulgata} and \emph{Nacella
deaureata} have been studied using Mg/Ca with mixed or contrary results.
In this study, we present elemental maps of various such species
together with stable oxygen isotope values for some of the specimens.
Our dataset also includes specimens that were previously unsuccessful in
providing significant correlations between δ\textsuperscript{18}O and
Mg/Ca ratios. By reassessing these previous specimens and including a
wider range of modern and archaeological samples from three patelloid
species (\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N.
magellanica}) we further add to the growing set of evidence for the
reliable use of Mg/Ca ratios to detect palaeotemperature change and
serve as a means to determine ontogenetic age and season of capture as
well as to reveal locations of interest within the growth record
(i.e.~annual temperature minima and maxima) for the targeted analysis
using δ\textsuperscript{18}O or clumped oxygen isotope analysis.
\end{abstract}

@@

\subsubsection{Modern specimens}\label{modern-specimens}

The analysed specimens, their origin and respective studies with
research background information can be found in \hyperref[Table_1]{Table
1}. Here we will briefly summarise their contextual information, which
can be accessed in more detail at the respective studies
\citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While those
studies also included other specimens, their accessibility or state of
preservation did not all lend them to be re-analysed.

\phantomsection\label{Table_1}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
\caption{Overview of the modern and archaeological limpet specimens
analysed in this study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
- Tierra del Fuego (AR) & ND-1016-3 & Stable oxygen and carbon isotope
analysis \\
& & & & ND-1016-4 & \\
& & \emph{N. magellanica} & & NM-1016-1 & \\
& & & & NM-1016-3 & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - Westray
(UK) & ORK-LT5 & Stable oxygen and carbon isotope analysis; Mg, Li, Sr,
Ca,. \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - Westray (UK) & QG1-7188-1 & Stable oxygen and carbon isotope
analysis \\
& & & & QG1-7189-2 & \\
& & & & QG2-1061-1 & \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
& This study & \emph{N. magellanica} & Heshkaia 35 in Moat Bay - Tierra
del Fuego (AR) & Heshkaia\_35-1 & \emph{none} \\
& & \emph{N. deaureata} & & Heshkaia\_35-2 & \\
& & \emph{N. deaureata} & & Heshkaia\_35-3 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-4 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-5 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-6 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-7 & \\
& & \emph{N. deaureata} & & Heshkaia\_35-8 & \\
& & \emph{N. magellanica} & & Heshkaia\_35-9 & \\
\end{longtable}

@@

@@

\includegraphics{Manuscript_files/figure-pdf/fig-SSTs-1.pdf}

@@

\caption{\label{fig-SSTs}SST values averaged from years. SST values were
sourced from \citet{Good2020-nl}}

@@

\subsubsection{Archaeological specimens}\label{archaeological-specimens}

In total, we studied 16 archaeological specimens. Of those, seven
\emph{P. vulgata} come from Quoygrew on Westray dating to 900 -- 1200
CE. These shells have previously been studied to characterise seasonal
temperature change during the Medieval Climate Anomaly and the Little
Ice Age \citep{Surge2012-ba}. Together they show a range of lifespans,
with some comparatively long for their size (QG1-7246-1, 12 years within
16 mm of growth) making them particularly susceptible to time-averaging.
That said, the majority are under 5 years old. These shells are an ideal
comparative dataset from an archaeological assemblage to compare to the
modern shell ORK\_LT5 \citep{Graniero2017-io}.

Nine additional shells of \emph{N. magellanica} (n=6) and \emph{N}.
\emph{deaureata} (n=3) have been selected for LIBS analysis only, to
provide a comparative dataset to the modern \emph{Nacella} sp. shells.
These have not been analysed for δ\textsuperscript{18}O-values, but
rather as a test to see whether patterns visible in the modern Mg/Ca
ratio distributions are also preserved and apparent in the
archaeological shells. They derive from the site of Heshkaia 35 situated
within Moat Bay, around 30 km east of where the modern specimens in our
study were collected \citep{Zangrando2014-qi, Zangrando2021-qg}. With
radiocarbon dates putting the shells into the Late Holocene (1300 --
1450 CE) they are of similar age --- if somewhat younger --- as the
assemblage from Quoygrew.

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Nac_iso-1.pdf}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_iso-1.pdf}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_LIBS-1.pdf}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Comp-1.pdf}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Pat_Corr-1.pdf}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Nac_LIBS-1.pdf}

@@

Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
shells were much lower (0.05--0.30) than those seen in other patelloid
species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in )
using the same emission lines. This is particularly visible in
Figure~\ref{fig-Nac_LIBS} c, where Mg/Ca intensity ratios range from
\textasciitilde0.5--0.7 in \citep{Hausmann2023-ih}). While the intensity
ratio depends chiefly on the chosen emission lines to calculate the
proportion of the chosen Magnesium to the Calcium peak, the low ratios
still indicate lower concentrations than seen elsewhere, or seen only in
aragonitic parts of e.g.~\emph{Patella} shells.

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Comp-1.pdf}

@@

\includegraphics{Manuscript_files/figure-pdf/fig-Nac_Corr-1.pdf}

@@

\section{Discussion}\label{discussion}

\subsection{Other correlations}\label{other-correlations}

{\centering \includegraphics{Manuscript_files/figure-pdf/Ferguson Data-1.pdf}

@@

\caption{Correlation graphs for Ferguson et al.~specimens}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2000}}@{}}
\caption{Table 2: Overview of comparative correlations
\{\#tab:correlations\}}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Correlation R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Correlation R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{P. depressa} & Northern Spain & LAN541 & 0.87 &
\citep{Garcia-Escarzaga2021-ij} \\
& & LAN545 & 0.86 & \\
& & LAN554 & 0.78 & \\
& & LAN559 & 0.82 & \\
\emph{P. caerulea} & Croatia & ISTPC1 & 0.9 & \citep{Hausmann2019-fi} \\
& & ISTPC2 & 0.84 & \\
& Crete & AF1911A & 0.91\footnote{SST only, no other geochemical data
  available} & \\
& & AF3003A & 0.92\footnote{SST only, no geochemical data available}
& \\
& Israel & AKKPC2 & 0.96 & \\
& & AKKPC3 & 0.89 & \\
& & FRMPC1 & 0.84 & \\
& & FRMPC2 & 0.96 & \\
& Libya & MO31A & 0.83 & \\
& & MP64A & 0.33 & \\
& & MP67A & 0.96 & \\
& & MP68A & 0.81 & \\
& Malta & MA10 & 0.82 & \\
& Tunisia & TUNPC1 & 0.81 & \\
& & TUNPC2 & 0.78 & \\
& Turkey & ANTPC1 & 0.95 & \\
& & ANTPC2 & 0.93 & \\
& & KIZPC1 & 0.94 & \\
& & KIZPC2 & 0.86 & \\
\emph{P. rustica} & Gibraltar & JL1 & 0.02 & \citep{Ferguson2011-zl} \\
& & JL2 & 0.8 (0.79) & \\
\emph{P. caerulea} & Gibraltar & JM00 & 0.69 (0.79) & \\
& & JM30 & 0.83 (0.79) & \\
\emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
\citep{Graniero2017-io} and this study \\
& & & & \\
& & & & \\
& & & & \\
\end{longtable}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Interspecies comparisons of Mg/Ca ratios in limpet shells},
  pdfauthor={Niklas Hausmann; Donna Surge; Ivan Briz-Godino},
  pdfkeywords={Sclerochronology, Limpets, Elemental Ratio, Mg/Ca},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\author[3]{Ivan Briz-Godino%
%
}

\affiliation[1]{organization={Leibniz Zentrum für
Archäologie},addressline={Ludwig-Lindenschmit-Forum
1},city={Mainz},country={Germany},countrysep={,},postcode={55116},postcodesep={}}
\affiliation[2]{organization={University of North
Carolina},addressline={104 South Road, 225 Geology
Building},city={Chapel Hill,
NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
\affiliation[3]{organization={},country={Spain},countrysep={,},postcodesep={}}

@@

\begin{abstract}
This study provides a short reassessment of the use of Magnesium to
Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea
surface temperatures. While \emph{Patella vulgata} along the Spanish
shoreline has since then repeatedly produced reliable correlations
between sea surface temperature and Mg/Ca ratios, this is not the case
for other patelloid species. \emph{Patella vulgata} and \emph{Nacella
deaureata} have been studied using Mg/Ca with mixed or contrary results.
In this study, we present elemental maps of various such species
together with stable oxygen isotope values. Our dataset also includes
specimens that were previously unsuccessful in providing significant
correlations between δ\textsuperscript{18}O and Mg/Ca ratios. By
reassessing these previous specimens and including a wider range of
modern and archaeological samples (n=12) from three patelloid species
(\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N. magellanica}) we
further add to the growing set of evidence for the reliable use of Mg/Ca
ratios to detect palaeotemperature change and serve as a means to
determine ontogenetic age and season of capture as well as to reveal
locations of interest within the growth record (i.e.~annual temperature
minima and maxima) for the targeted analysis using
δ\textsuperscript{18}O or clumped oxygen isotope analysis.
\end{abstract}

@@

The analysed specimens, their origin and respective studies with
research background information can be found in
Table\hyperref[Table_1]{1}. Here we will briefly summarise their
contextual information, which can be accessed in more detail at the
respective studies
\citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While those
studies also included other specimens, their accessibility or state of
preservation did not all lend them to be re-analysed.

@@

\phantomsection\label{Table_1}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1667}}@{}}
\caption{Overview of the modern and archaeological limpet specimens
analysed in this study.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Context
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Location
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Sample ID
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Previous analyses
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Modern & \citep{Nicastro2020-ih} & \emph{N. deaureata} & Cambaceres Bay
- Tierra del Fuego (AR) & ND-1016-3 & Stable oxygen and carbon isotope
analysis \\
& & & & ND-1016-4 & \\
& & \emph{N. magellanica} & & NM-1016-1 & \\
& & & & NM-1016-3 & \\
& \citep{Graniero2017-io} & \emph{P. vulgata} & Rack Wick Bay - Westray
(UK) & ORK-LT5 & Stable oxygen and carbon isotope analysis; Mg, Li, Sr,
Ca,. \\
Archaeological & \citep{Surge2012-ba} & \emph{P. vulgata} & Rack Wick
Bay - Westray (UK) & QG1-7188-1 & Stable oxygen and carbon isotope
analysis \\
& & & & QG1-7189-2 & \\
& & & & QG2-1061-1 & \\
& & & & QG2-1064-1 & \\
& & & & QG2-7180-1 & \\
& & & & QG2-7180-2 & \\
& & & & QG1-7246-1 & \\
\end{longtable}

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-SSTs-1.pdf}

@@

\caption{\label{fig-SSTs}Monthly Sea Surface Temperature values averaged
from years. SST values were sourced from \citet{Good2020-nl} .}

@@

In total, we studied 7 archaeological specimens of \emph{P. vulgata}
coming from Quoygrew on Westray dating to 900 -- 1200 CE. These shells
have previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly and the Little Ice Age
\citep{Surge2012-ba}. Together they show a range of lifespans, with some
comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
growth) making them particularly susceptible to time-averaging. That
said, the majority are under 5 years old. These shells are an ideal
comparative dataset from an archaeological assemblage to compare to the
modern shell ORK\_LT5 \citep{Graniero2017-io}.

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Nac_iso-1.pdf}

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Pat_iso-1.pdf}

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Pat_LIBS-1.pdf}

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Pat_Comp-1.pdf}

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Pat_Corr-1.pdf}

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Nac_LIBS-1.pdf}

@@

Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
shells were much lower (0.05--0.30) than those seen in other patelloid
species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in )
using the same emission lines. This is particularly visible in
Figure~\ref{fig-Nac_LIBS} c, where Mg/Ca intensity ratios range from
\textasciitilde0.5--0.7 in \citep{Hausmann2023-ih}). While the intensity
ratio depends chiefly on the chosen emission lines to calculate the
proportion of the chosen Magnesium to the Calcium peak, the low ratios
still indicate lower concentrations than seen elsewhere, or seen only in
aragonitic parts of e.g.~\emph{Patella} shells. Similar observations
have been made before \citep{Graniero2017-io}.

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Nac_Comp-1.pdf}

@@

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Nac_Corr-1.pdf}

@@

\section{Discussion}\label{Discussion}

The Mg/Ca ratios from the 2D elemental imaging present positive results
on SST being the main factor behind Mg/Ca concentration in three
Atlantic limpet species. Annually repeating patterns of increasing
(spring/summer) and decreasing (autum/winter) SST were also visible in
the line scans, albeit less straightforward than in the 2D-data.
Nevertheless, the resulting R\textsuperscript{2}-values for each species
were high (\emph{P. vulgata mean R\textsuperscript{2}}=0.89, \emph{N.
deaureata mean R\textsuperscript{2}}=0.86\emph{, N. magellanica mean
R\textsuperscript{2}}=0.88). These results are similar to previously
published data on \emph{Patella} sp. shells (Table
\hyperref[Table_2]{2}). R\textsuperscript{2} values of \emph{P.
depressa} lie between 0.78 and 0.87 \citep{Garcia-Escarzaga2021-ij}, for
\emph{P. caerulea} from the Mediterranean between 0.78 and 0.96 ---
although with one outlier of 0.33, which had increased Mg/Ca ratios in
increments with rich organic contents \citep{Hausmann2019-fi}.

\emph{P. rustica} has two studied specimens (JL1 and JL2) with
R\textsuperscript{2}-values of 0.02 and 0.8, respectively
\citep{Ferguson2011-zl} (Figure~\ref{fig-Ferg}). Together with the data
from two \emph{P. caerulea} shells (JM00 and JM30 with
R\textsuperscript{2}-values of 0.69 and 0.83), the data from JL2 was
used determine a general Mg/Ca-SST relationship with an
R\textsuperscript{2}-value of 0.79. JL1 was disregarded in that
calculation, as well as the second year of growth in JM00, which
produced anomalous data. While no additional information about those
specimens is available, it seems likely that the Mg/Ca ratios from JM00,
which were interpreted as an ontogenetic trend, were potentially
influenced by the same intra-increment heterogeneity we see in some of
our shell specimens as well. Since these shells were analysed by
homogenising the carbonate powder of the m+2 layer and the more exterior
m+3 layer, these heterogeneities could potentially increased the Mg/Ca
ratio unpredictably, leading to what might look like an ontogenetic
trend in a linear dataset. Whether these anomalies affected JL1 in the
same way, albeit stronger, is not clear but it might well be possible.

\phantomsection\label{Table_2}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1918}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1918}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1918}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1918}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2329}}@{}}
\caption{Table showing correlations of Mg/Ca and δ\textsuperscript{18}O
found in other studies. * indicates that only SST data and no other
geochemical data was used. ** these samples were used to determine a
shared R\textsuperscript{2}-value of 0.79.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Species
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Locality
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Specimen
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Study
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\emph{P. depressa} & Northern Spain & LAN541 & 0.87 &
\citep{Garcia-Escarzaga2021-ij} \\
& & LAN545 & 0.86 & \\
& & LAN554 & 0.78 & \\
& & LAN559 & 0.82 & \\
\emph{P. caerulea} & Croatia & ISTPC1 & 0.9 & \citep{Hausmann2019-fi} \\
& & ISTPC2 & 0.84 & \\
& Crete & AF1911A & 0.91 & \\
& & AF3003A & 0.92 & \\
& Israel & AKKPC2 & 0.96 & \\
& & AKKPC3 & 0.89 & \\
& & FRMPC1 & 0.84 & \\
& & FRMPC2 & 0.96 & \\
& Libya & MO31A & 0.83 & \\
& & MP64A & 0.33 & \\
& & MP67A & 0.96 & \\
& & MP68A & 0.81 & \\
& Malta & MA10 & 0.82 & \\
& Tunisia & TUNPC1 & 0.81 & \\
& & TUNPC2 & 0.78 & \\
& Turkey & ANTPC1 & 0.95 & \\
& & ANTPC2 & 0.93 & \\
& & KIZPC1 & 0.94 & \\
& & KIZPC2 & 0.86 & \\
\emph{P. rustica} & Gibraltar & JL1 & 0.02 & \citep{Ferguson2011-zl} \\
& & JL2 & 0.8** & \\
\emph{P. caerulea} & Gibraltar & JM00 & 0.69** & \\
& & JM30 & 0.83** & \\
\emph{P. vulgata} & Orkney & ORK-LT5 & not reported, here 0.88 &
\citep{Graniero2017-io} and this study \\
\end{longtable}

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-Ferg-1.pdf}

@@

\caption{\label{fig-Ferg}Correlation graphs for Ferguson et al., 2011
specimens.}

@@

Another shell whose previous results we aim to explain further is
ORK-LT5, which has previously been shown to have no substantial
correlation with δ\textsuperscript{18}O or SST in implication
\citep{Graniero2017-io}. Different to other specimens of this study, we
were able to resample this specimen and getting Mg/Ca data on a slightly
higher resolution (20--30 µm spots in a 100 µm raster compared to 50 µm
spots spaced at 150--300 µm) and using 2 dimensions, as well as using a
continuous high-resolution line-scan at (20--30 µm spots spaced at 10
µm) (Figure~\ref{fig-ORK_sub}). These additional data revealed
previously missed summer peaks at around 2--3 mm distance to the shell
edge (grey box in Figure~\ref{fig-ORK_sub}) and revealed additional
growth at the very edge (dashed box in Figure~\ref{fig-ORK_sub})
mirroring the change towards warmer temperatures at the time of
collection (August). Including these missing parts lead to a better
alignment of both records and a resulting R\textsuperscript{2}-value of
0.88.

\includegraphics{Manuscript_with_Heshkaia_files/figure-pdf/fig-ORK_sub-1.pdf}

\caption{\label{fig-ORK_sub}Comparison of previous and new Mg/Ca data
for ORK-LT5 explaining the previously unsuccessful correlation of Mg/Ca
and δ\textsuperscript{18}O-values}

While we were not able to resample the \emph{Nacella} sp. specimens in
Graniero \emph{et al.} \citeyearpar{Graniero2017-io}, we were able to
source other modern and archaeological specimens of the same species
from Tierra del Fuego, which can serve as an indicator. Similar to other
patelloid shells, the results of the \emph{Nacella} specimens in our
study were successful with high R\textsuperscript{2}-values (0.76--0.95)
and suggest Mg/Ca as a reliable proxy for SST (Figure~\ref{fig-Nac_Comp}
and supplementaries). Why these results are different from the studied
\emph{Nacella} sp. specimens in Graniero et al.
\citeyearpar{Graniero2017-io} is not obvious, but the occasional
alignment of the Mg/Ca and δ\textsuperscript{18}O records of their
shells (Figure 6 in \citep{Graniero2017-io} ) indicate that there is
some relationship that --- potentially due to the low sampling
resolution or intra-increment heterogeneity --- fell apart and lead to
an overall result of an uncorrelated relationship between Mg/Ca and SST.

\section{Conclusion}\label{conclusion}

This study presents high-resolution and 2D information on Mg/Ca
concentrations in modern and archaeological patelloid shells and their
relation to sea surface temperature. The results for \emph{Patella
vulgata}, \emph{Nacella deaureata}, and \emph{Nacella magellanica} were
consistently promising with mean R\textsuperscript{2}-values of 0.89,
0.86, and 0.88 respectively.

LIBS elemental imaging provided some insight into elemental anomalies
found particulary in the external parts (m+3 layer) of the shell, which
introduce intra-incremental heterogeneity and which cannot be reconciled
with modern SST changes or SST changes recorded in the
δ\textsuperscript{18}O of the same shells. Our 2D approach allowed us to
avoid these areas of intra-increment heterogeneity and focus on areas
that are predominantly controlled by changes in SST.

These results are particularly promising for future analyses of
patelloid shells for the study of season of collection in archaeological
sites as well as the reconstruction of sub-annual climatic conditions in
the past. Shells of the species studied here are widely available along
the European Atlantic façade and the Argentinian coastline and make up a
large part of marine mollusc remains in archaeological sites of those
regions. Improving their analysis in terms of cost and time efficiency
through the benefits of LIBS, which can be used in this context as a
screening tool for targeted δ\textsuperscript{18}O analysis focusing on
annual minima and maxima revealed in the Mg/Ca record, allows
researchers to increase sample numbers and produce larger and more
in-depth climatic studies than previously possible.
